Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2021-05-11 2021/emnlp/auto_cl_mdmt.tex
review section 3

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions \citep{Moore10selection,Axelrod11domain}. Finding the optimal balance of training data is however a challenging task due, for instance, to the similarity between domains/languages, but also due to regularization effects of out-of-domain data. It may also be suboptimal, as some target domains or languages might be easier to train than others. For example, improving the performance of the MT system in one domain will often hurt that of another \citep{Vanderwees17dynamic, Britz17mixing}, and improving model generalization across all domains \citep{koehn18findings} may not promise optimization of a particular domain.

@@

We conventionally define a domain $d$ as a distribution $\mathcal{D}_d(x)$ over some feature space $\mathcal{X}$ that is shared across domains \citep{Pan10asurvey}: in machine translation, $\mathcal{X}$ is the representation space for input sentences; each domain corresponds to a specific source of data, and may differ from the other data sources in terms of textual genre, thematic content \cite{Chen16guided,Zhang16topicinformed}, register \cite{Sennrich16politeness}, style \cite{Niu18multitask}, etc. Translation in domain $d$ is formalized by a translation function $h_d(y|x)$ pairing sentences in a source language with sentences in a target language $y \in \mathcal{Y}$. $h_d$ is usually assumed to be deterministic (hence $y = h_d(x)$), but might differ from one domain to the other.

It is usual in MT to opportunistically collect training samples from several domains, which means that training instances are distributed according to the mixture $\mathcal{D}^s$ such that $\mathcal{D}^s(x) = \sum_{d=1}^{n_d} \lambda^{s}(d) \mathcal{D}_d(x)$, where $\{\lambda^{s}(d), d=1 \dots n_d\}$ are the corresponding mixture weights satisfying $\sum_d \lambda^{s}(d)=1$.

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unsupervised domain adaptation); in multi-domain adaptation finally, the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of fine-tuning.
These situations are illustrated in Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

@@

A more practical standard approach to supervised DA is \emph{fine-tuning} \cite{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyTodo{How to measure this?} \revision{Answer: The graphic of \system{Mixed-0} shows that with the same amount of data(batches) the convergences of the domains vary significantly}

@@

\section{Multi-Domain Automated Curriculum } \label{sec:mdac}
\subsection{Basic principles}
Assuming training data from $n_d$ domains $d_1 \dots d_{n_d}$, we denote the size of the training corpus from domain $d$ as  $N^{s}_d$, and $N^{s} = \sum_d N^{s}_d$ is the total number of training samples. We use $\widehat{\mathcal{D}^l_d}$ and $\widehat{\mathcal{D}^t _d}$ to denote the empirical train and test distributions for domain $d$, and $\widehat{\mathcal{D}^{u}}(x;\lambda^{u}) = \sum_{d} \lambda^{u}(d) \widehat{\mathcal{D}^{u}_d}(x)$ for $u\in\{l,t\}$. In our setting,  $\vlambda^t$, and hence $\widehat{\mathcal{D}^t}(x;\vlambda^t)$ are fixed and predefined, approximated with an equivalent number of development corpora.

Our method MDAC constructs an automatically adapting distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution via REINFORCE algorithm \citep{Williams92simple}.

\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{1} * \displaystyle{\mathop{\sum}_{d=1}^{n_d}} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi} \\
\end{align*}
\begingroup
\allowdisplaybreaks
With the following reward computation:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t)
\end{align*}
Where 
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& Update\big(\theta_{t+i-1},[x^i_j,y^i_j]_{j=1}^N\big) \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d)
\end{array}
\end{equation}
\endgroup
%\theta_{t+i-1} - \mathbf{lr_2} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta}
$N$ is the size of a batch; $\mathbf{lr}_{1}$ is the learning rate of the sampling distribution; $\mathbf{lr_2}$ is the learning rate of the NMT system; $l(\theta,x,y)$ is the loss of the NMT model on the sample $(x,y)$; $J^t(\theta,\lambda_t)$ is the weighted loss computed over $n_d$ dev-sets corresponding to $n_d$ given domains.

To compute the reward of using data from domain $d$ to train the model, $R(d)$, we simulate $k$ training steps from the current checkpoint by training the model with k batches sampled from $D^l(d)$ and compute the gain of the weighted dev-loss. The reward of our method is inspired by the target prediction gain in \cite{Graves17automated}. However, \cite{Graves17automated} used accumulated gains from the past as reward while we predict the utility of each domain for improving the competence of the system at the current state by simulating the training with the data of the domain. Furthermore, the choice of the parameterization of the sampling distribution in our method is different from one in \cite{Graves17automated}, although this choice is empirical.

The work of \cite{Wang20balancing} based on the Bi-level Optimization framework, which aims to find an optimal static distribution $\lambda^{l}$ that will result in the best NMT model regarding a given target dev set at the end of the training, derives also a similar form of update for $\psi$. However, their reward is the cosine similarity between the gradient computed with the training data and the gradient computed with the dev set.

\subsection{MDAC for (multi) domain adaptation}
This setting is pretty general and can in principle accommodate the variety of situations mentioned  above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments we would like to better assess the actual potential of MDAC is these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is MDAC a viable alternative to conventional fine-tuning? In particular does it enable to better take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform unsupervised (multi-) domain adaptation? \fyDone{TBContinued}
\end{itemize}

These questions are further explored in Section~\ref{sec:results}. We now turn to our experimental conditions.

>>>>

This problem is often approached by \emph{static} instance selection or re-weighting strategies, where the available training data is used in proportion to its relevance for the chosen test conditions \citep{Moore10selection,Axelrod11domain}. Finding the optimal balance of training data is however a challenging task due, for instance, to the similarity between domains/languages, but also due to regularization effects of out-of-domain data. It may also be suboptimal, as some target domains or languages might be easier to train than others. For example, improving the performance of the MT system in one domain will often hurt that of another \citep{Vanderwees17dynamic, Britz17mixing}, and improving model generalization across all domains \citep{koehn18findings} may not achieve optimily for any particular domain.

@@

We conventionally define a domain $d$ as a distribution $\mathcal{D}_d(x)$ over some feature space $\mathcal{X}$ that is shared across domains \citep{Pan10asurvey}: in machine translation, $\mathcal{X}$ is the representation space for input sentences; each domain corresponds to a specific source of data, and may differ from the other data sources in terms of textual genre, thematic content \citep{Chen16guided,Zhang16topicinformed}, register \cite{Sennrich16politeness}, style \cite{Niu18multitask}, etc. Translation in domain $d$ is formalized by a translation function $h_d(y|x)$ pairing sentences in a source language with sentences in a target language $y \in \mathcal{Y}$. $h_d$ is usually assumed to be deterministic (hence $y = h_d(x)$), but might differ from one domain to the other.

It is usual in MT to opportunistically collect training samples from several domains, which means that training instances are distributed according to a mixture $\mathcal{D}^s$ such that $\mathcal{D}^s(x) = \sum_{d=1}^{n_d} \lambda^{s}(d) \mathcal{D}_d(x)$, with $\{\lambda^{s}(d), d=1 \dots n_d\}$ the mixture weights satisfying $\sum_d \lambda^{s}(d)=1$.

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unsupervised domain adaptation); in multi-domain adaptation finally, the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of pure domain adaptation.
These various settings are illustrated Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

@@

A more practical standard approach to supervised DA is \emph{finetuning} \cite{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyTodo{How to measure this?} \revision{Answer: The graphic of \system{Mixed-0} shows that with the same amount of data(batches) the convergences of the domains vary significantly}

@@

\section{Multi-Domain Automated Curriculum } \label{sec:mdac}
\subsection{Basic principles}
Assuming training data in each of the $n_d$ domains $d_1 \dots d_{n_d}$, we denote the size of the training corpus from domain $d$ as $N^{s}_d$, and $N^{s} = \sum_d N^{s}_d$ is the total number of training samples. We use $\widehat{\mathcal{D}^l_d}$ and $\widehat{\mathcal{D}^t _d}$ to denote the empirical train and test distributions for domain $d$, and $\widehat{\mathcal{D}^{u}}(x;\lambda^{u}) = \sum_{d} \lambda^{u}(d) \widehat{\mathcal{D}^{u}_d}(x)$ for $u\in\{l,t\}$. In our setting,  $\vlambda^t$, and hence $\widehat{\mathcal{D}^t}(x;\vlambda^t)$ are fixed and predefined, approximated with an equivalent number of development corpora.

% Our method
MDAC constructs an adaptative training distribution $\lambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\lambda^{l}$ by a differentiable function $\lambda^l(\psi)$. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\lambda^{l}(\psi_t)$. After one learning session, we update the data distribution using the REINFORCE algorithm of \citet{Williams92simple}. The evolution of $\psi$ is thus defined by:
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{1} * \displaystyle{\mathop{\sum}_{d=1}^{n_d}} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi}, \\
\end{align*}
\begingroup
\allowdisplaybreaks
where reward $R(d)$ is computed as:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t),
\end{align*}
and where we also define:
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& Update\big(\theta_{t+i-1},[x^i_j,y^i_j]_{j=1}^N\big) \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\lambda_t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d).
\end{array}
\end{equation}
\endgroup
% \theta_{t+i-1} - \mathbf{lr_2} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta}
In these equations, $N$ denotes the size of a batch; $\mathbf{lr}_{1}$ is the learning rate of the sampling distribution; $\mathbf{lr_2}$ is the learning rate of the NMT system; $l(\theta,x,y)$ is the loss of the NMT model on sample $(x,y)$; $J^t(\theta,\lambda_t)$ is the weighted loss aggregated over $n_d$ dev-sets corresponding to the $n_d$ domains.

To compute the reward $R(d)$ of using data from domain $d$ to train the model, we simulate $k$ training steps from current checkpoint by training the model with $k$ batches sampled from $D^l(d)$ and compute the gain of the weighted dev-loss. This computation is inspired by the target prediction gain in \citep{Graves17automated}. However, \cite{Graves17automated} used accumulated gains from the past as reward, while we predict the usefulness of each domain for improving the future performance of the system given its current state. This is achieved by simulating a round of training with only the data from one domain. Furthermore, the choice of the parameterization of the sampling distribution in our method differs from that of \citet{Graves17automated}, although this choice is empirical.

The work of \cite{Wang20balancing} is also related: it is based on the Bi-level Optimization framework, which aims to find an optimal static distribution $\lambda^{l}$ that will result in the best NMT model with respect to a given target dev set at the end of training. These authors also derive a similar form of update for $\psi$. However, their reward is the cosine similarity between the gradient computed with the training data from one domain and the gradient computed with the dev set. We compare this approach with ours in the experiment section.

\subsection{MDAC for (multi) domain adaptation}
The setting developed in previous sections is quite general and can in principle accommodate the variety of situations mentioned above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments, we would like to better assess the true potential of MDAC is these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is MDAC a viable alternative to conventional finetuning? In particular, does it enable to better take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform unsupervised (multi-) domain adaptation? \fyDone{TBContinued}
\end{itemize}
These questions are further explored in section~\ref{sec:results}. We now turn to our experimental conditions.
